---
title: "Edge Dryspot Detection in Human Hippocampus Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Edge Dryspot Detection in Human Hippocampus Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
pkgload::load_all("..", export_all = TRUE)
knitr::opts_chunk$set(cache = FALSE)
library(SpatialDryArtifacts)
library(ExperimentHub)
library(ggplot2)
library(gridExtra)
```

Introduction
This vignette demonstrates how to use SpatialDryArtifacts to detect and classify edge dryspots in spatial transcriptomics data using human hippocampus samples.

Loading Data
We'll use the human hippocampus dataset from ExperimentHub:

library(SpatialDryArtifacts)
library(ExperimentHub)
library(ggplot2)
library(gridExtra)

```{r loadingData}
# Load hippocampus data
eh <- ExperimentHub()
myfiles <- query(eh, "humanHippocampus2024")
spatial_hpc_spe <- myfiles[["EH9605"]]
cat("Total samples:", length(unique(spatial_hpc_spe$sample_id)), "\n")
cat("Total spots:", ncol(spatial_hpc_spe), "\n")
```
Edge Dryspot Detection
We'll demonstrate detection using balanced parameters that provide good sensitivity while avoiding over-detection:

```{r detectingSpot, cache=FALSE}
# Clear R session state to prevent cumulative raster errors
gc()  # Force garbage collection
if(requireNamespace("raster", quietly = TRUE)) {
  raster::removeTmpFiles(h=0)  # Clear raster temporary files
}

# Apply edge dryspot detection to all samples
spe_detected <- detectEdgeDryspots(
  spatial_hpc_spe,
  qc_metric = "sum_gene",
  samples = "sample_id",
  mad_threshold = 1.8,
  edge_threshold = 0.2,
  min_cluster_size = 3,
  name = "edge_dryspot"
)

# Apply enhanced classification
spe_classified <- classifyEdgeDryspots(spe_detected)
```

Visualization: QC Metrics and Detection Results
We'll create a comprehensive visualization showing QC metrics and detection results for all samples:
```{r plot}
# Get all unique samples
all_samples <- unique(spe_classified$sample_id)

# Create plotting function for each sample
create_sample_plots <- function(sample_id, spe_data) {
  
  # Extract single sample
  spe_sample <- spe_data[, spe_data$sample_id == sample_id]
  coords <- spatialCoords(spe_sample)
  in_tissue <- spe_sample$in_tissue
  
  plot_data <- data.frame(
    x = coords[in_tissue, 1],
    y = coords[in_tissue, 2],
    sum_umi = spe_sample$sum_umi[in_tissue],
    sum_gene = spe_sample$sum_gene[in_tissue],
    expr_chrM_ratio = ifelse(is.na(spe_sample$expr_chrM_ratio[in_tissue]), 0, 
                             spe_sample$expr_chrM_ratio[in_tissue]),
    edge_flagged = spe_sample$edge_dryspot_edge[in_tissue]
  )
  
  # Base theme
  base_theme <- theme_void() + 
    theme(plot.title = element_text(size = 8, hjust = 0.5),
          legend.position = "none")
  
  # Create 4 plots per sample
  p1 <- ggplot(plot_data, aes(x = x, y = y, color = log10(sum_umi + 1))) +
    geom_point(size = 0.3, alpha = 0.7) +
    scale_color_viridis_c(name = "log10(UMI)") +
    ggtitle(paste(sample_id, "- UMI")) +
    base_theme + coord_fixed()
  
  p2 <- ggplot(plot_data, aes(x = x, y = y, color = log10(sum_gene + 1))) +
    geom_point(size = 0.3, alpha = 0.7) +
    scale_color_viridis_c(name = "log10(Gene)", option = "plasma") +
    ggtitle(paste(sample_id, "- Genes")) +
    base_theme + coord_fixed()
  
  p3 <- ggplot(plot_data, aes(x = x, y = y, color = expr_chrM_ratio)) +
    geom_point(size = 0.3, alpha = 0.7) +
    scale_color_viridis_c(name = "Mito Ratio", option = "inferno") +
    ggtitle(paste(sample_id, "- Mito")) +
    base_theme + coord_fixed()
  
  p4 <- ggplot(plot_data, aes(x = x, y = y, color = edge_flagged)) +
    geom_point(size = 0.3, alpha = 0.7) +
    scale_color_manual(values = c("FALSE" = "lightgray", "TRUE" = "red"), 
                       name = "Edge") +
    ggtitle(paste(sample_id, "- Flagged")) +
    base_theme + coord_fixed()
  
  return(list(p1, p2, p3, p4))
}

# Generate plots for all samples
cat("Generating plots for", length(all_samples), "samples...\n")
all_plots <- list()

for (sample_id in all_samples) {
  sample_plots <- create_sample_plots(sample_id, spe_classified)
  all_plots <- append(all_plots, sample_plots)
}

# Create combined visualization (36 rows x 4 columns)
combined_plot <- do.call(gridExtra::grid.arrange, c(
  all_plots,
  ncol = 4,  # 4 columns: UMI, Genes, Mito, Flagged
  nrow = length(all_samples),  # One row per sample
  top = "Human Hippocampus Edge Dryspot Detection - All Samples"
))

print(combined_plot)
```

Classification Summary
Let's examine the enhanced classification system:
```{r classification}
# Create classification summary table
classification_summary <- table(spe_classified$edge_dryspot_enhanced)
print("Enhanced Classification Summary:")
print(classification_summary)

# Calculate percentages
classification_pct <- round(100 * classification_summary / sum(classification_summary), 2)
print("Classification Percentages:")
print(classification_pct)

# Create summary data frame
summary_df <- data.frame(
  Classification = names(classification_summary),
  Count = as.numeric(classification_summary),
  Percentage = as.numeric(classification_pct)
)

print(summary_df)

# Sample-wise summary
sample_summary <- aggregate(
  spe_classified$edge_dryspot_edge, 
  by = list(Sample = spe_classified$sample_id), 
  FUN = function(x) sum(x, na.rm = TRUE)
)
names(sample_summary)[2] <- "Edge_Spots_Detected"

sample_summary$Total_Spots <- aggregate(
  spe_classified$in_tissue, 
  by = list(spe_classified$sample_id), 
  FUN = sum
)$x

sample_summary$Detection_Percentage <- round(
  100 * sample_summary$Edge_Spots_Detected / sample_summary$Total_Spots, 2
)

print("Sample-wise Detection Summary:")
print(sample_summary)
```
Quality Control Validation
Finally, let's validate that flagged spots have lower quality metrics:

```{r validation}
# Create QC comparison data
in_tissue_data <- spe_classified[, spe_classified$in_tissue]

qc_data <- data.frame(
  sum_umi = in_tissue_data$sum_umi,
  sum_gene = in_tissue_data$sum_gene,
  expr_chrM_ratio = ifelse(is.na(in_tissue_data$expr_chrM_ratio), 0, 
                           in_tissue_data$expr_chrM_ratio),
  flagged = in_tissue_data$edge_dryspot_edge
)

# Calculate median differences
flagged_umi <- median(qc_data$sum_umi[qc_data$flagged])
nonflagged_umi <- median(qc_data$sum_umi[!qc_data$flagged])

flagged_gene <- median(qc_data$sum_gene[qc_data$flagged])
nonflagged_gene <- median(qc_data$sum_gene[!qc_data$flagged])

cat("QC Validation Results:\n")
cat("Flagged spots - Median UMI:", flagged_umi, "\n")
cat("Non-flagged spots - Median UMI:", nonflagged_umi, "\n")
cat("UMI difference:", nonflagged_umi - flagged_umi, "\n\n")

cat("Flagged spots - Median Genes:", flagged_gene, "\n")
cat("Non-flagged spots - Median Genes:", nonflagged_gene, "\n")
cat("Gene difference:", nonflagged_gene - flagged_gene, "\n")

# Create validation boxplot
qc_data$flag_status <- ifelse(qc_data$flagged, "Flagged", "Non-Flagged")

validation_plot <- ggplot(qc_data, aes(x = flag_status, y = log10(sum_umi + 1))) +
  geom_boxplot(fill = c("lightcoral", "lightblue")) +
  labs(title = "QC Validation: UMI Counts in Flagged vs Non-Flagged Spots",
       x = "Flag Status", y = "log10(UMI Count)") +
  theme_minimal()

print(validation_plot)
```

Conclusion
This vignette demonstrates:

Edge dryspot detection using balanced parameters on human hippocampus data
Comprehensive visualization showing QC metrics and detection results across all samples
Enhanced classification system with mechanistically-informed categories
Quality control validation confirming that flagged spots have lower quality metrics

The detection successfully identifies problematic spots while preserving high-quality tissue for downstream analysis.
